Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex
a few more bits

\krt{several refs seem missing}

@@

It was once a major undertaking to collect data sufficient to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
\jk{Citation?}
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
for instance, might have 500K individuals genotyped at 10M loci,
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Furthermore, even larger datasets are being collected,
and millions of whole genomes will soon be available. In this case the
genotype matrix simply cannot fit into memory, and complex distributed
computations are required to calculate population genetic statistics.

@@

\jk{These two intro paragraphs aren't working for me. There's too many
parenthetical comments, mixing high-level overview with getting into
the weeds. Does the first para actually help? Perhaps we're just confusing
people by talking about the AFS here?}
Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

% 2-3 sentances of intro before this?
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that edge.
Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

We now begin to formalize these notions with definitions.
In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
and the trees at each position along the genome have vertexes labeled by these nodes.
A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
that appear in the trees at every point along the genome.
The other (non-sample) nodes may not appear in every tree,
as for instance if a portion of an ancestor's genome was not inherited by any of the samples.

@@

\plr{not sure if we need this section}
\jk{Compress and move to the discussion?}

The statistics defined here are all \emph{additive} along the genome:
if we have computed a statistic in equal-size windows along a chromsome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
Multilocus statistics that measure linkage disequilibrium or haplotype lengths
do not fall in this category.
We have also defined site statistics to be additive across \emph{alleles},
to allow the definitions to apply to sites with more than two alleles.
This decision has some nice properties, but is not the only reasonable choice.

@@

\paragraph{Future work}
In this paper,
we have considered only so-called additive, single site statistics.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites
\citep[as used by][]{hudson2001twolocus},
such as linkage disequlibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

>>>>

@@

It was once a major undertaking to collect data sufficient to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
\jk{Citation?}
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread},
or between individuals rather than populations \citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}.
Computation is beginning to be a major problem:
genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
for instance, might have 500K individuals genotyped at 10M loci,
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Furthermore, even larger datasets are being collected,
and millions of whole genomes will soon be available. In this case the
genotype matrix simply cannot fit into memory, and complex distributed
computations are required to calculate population genetic statistics.

@@

Before formal definitions, we will give an overview of the general framework.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operations in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that branch.
Any mutation falling on this branch would be inherited by 
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to that described by \citet{fu1995statistical}.
To connect the framework to familiar notions,
here is a reframing of \citeauthor{fu1995statistical}'s statistics.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

Before defining the statistics, we need a few definitions.
In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
and the trees at each position along the genome have vertexes labeled by these nodes.
A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
that appear in the trees at every point along the genome.
The other (non-sample) nodes may not appear in every tree,
as for instance if a portion of an ancestor's genome was not inherited by any of the samples.

@@

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Duality in practice?}

The duality between Site and Branch statistics depends crucially on two assumptions:
that each new mutation is distinguishable, as in the infinite-sites model,
and that mutation rates are constant along the genome.
Neither of these things are true, but are they good approximations?
Multiple mutations at the same site, causing backmutations,
are present in real datasets but are generally thought to be uncommon,
especially if hypervariable sites are excluded.
The effect of multiple mutations on Site statistics
can be bounded by a function of the proportion of sites at which these occur.
Large-scale variation in mutation rate (at, say, the megabase scale)
might plausibly have a stronger effect on the correspondence,
because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
then dividing the Site statistic by the Branch statistic
would give us an estimate of relative average mutation rate in windows along the genome.
Similarly, Branch statistics from well-calibrated tree sequences
would provide a way to assess variation in population genetic summaries along the genome
in a way independent of local mutation rate.

A final crucial assumption underlying this duality is that mutations can be treated
as indepenent of the trees themselves.
This is clearly not always true --- for instance,
a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
but seems likely to be approximately true.
\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
but more work needs to be done, particularly on models with large proprotions of sites under selection.

The statistics defined here are all \emph{additive} along the genome:
if we have computed a statistic in equal-size windows along a chromsome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
We have also defined site statistics to be additive across \emph{alleles},
to allow the definitions to apply to sites with more than two alleles.
This decision has some nice properties, but is not the only reasonable choice.

Multilocus statistics that measure linkage disequilibrium or haplotype lengths
do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites
\citep[as used by][]{hudson2001twolocus},
such as linkage disequlibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

%%%%%%%%%%%%%%%%%%
\subsection*{Dynamic graph algorithms}

The tree sequence describes trees that change as one moves along the genome,
and so is a special case of a \emph{dynamic graph}
\citep[also called a \emph{graph timeline}][]{lacki2013reachability}.
Much of the work on dynamic graphs is focused on connectivity,
e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
but development of parallel algorithms for more general operations on large dynamic graphs
is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.
